---
title: "scExploreRFinal: Single-Cell RNA-seq Exploration"
author: |
  Leya Alexander  
  Undergraduate Pre-Med Student, University of South Florida (USF)  
  LIS4370: Advanced R Programming (Dr. Alon Friedman)  
  Machine Learning Department, Moffitt Cancer Center (El Naqa Lab)
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scExploreRFinal Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Introduction

Single-cell RNA sequencing (scRNA-seq) allows scientists to investigate cell
biology at an unprecedented resolution compared to traditional bulk RNA
sequencing. Instead of homogenizing tissues and losing critical tumor
microenvironment architecture, scRNA-seq captures the transcriptomic profile of
individual cells.

As an undergraduate pre-med student at the University of South Florida gaining
research experience at the El Naqa Lab (Moffitt Cancer Center), I am actively
investigating clear cell renal cell carcinoma (ccRCC) through single-cell and
spatial transcriptomics.

My exposure to these advanced methodologies deeply inspired the creation of this
project. While not an official product of the El Naqa Lab, it was built for
LIS4370 (Advanced R Programming) as an educational toolkit that encapsulates
complex genomics inside straightforward R paradigms.

---

## 1. Loading the Dataset

The package contains a light-weight subset of the classic 10X Genomics
Peripheral Blood Mononuclear Cells (PBMC) dataset. Loading it assigns the S3
class `scData`, which unlocks the package's generic methods.

```{r load}
library(scExploreRFinal)

# Load the dataset and create an S3 scData object
my_data <- load_scrna()

# The S3 print generic gives a concise summary
print(my_data)
```

```{r summary}
# summary() reports cell-type counts and library sizes (total counts per cell)
summary(my_data)
```

---

## 2. PCA Visualization — Three Plotting Systems

Principal Component Analysis (PCA) compresses high-dimensional gene expression
down to 2 axes so that cells with similar transcriptomic profiles cluster
together. The package demonstrates all three major R plotting systems.

### Base R

```{r pca-base, fig.width=7, fig.height=5}
plot_pca(my_data, method = "base")
```

### Lattice

```{r pca-lattice, fig.width=7, fig.height=5}
plot_pca(my_data, method = "lattice")
```

### ggplot2

```{r pca-ggplot2, fig.width=7, fig.height=5}
plot_pca(my_data, method = "ggplot2")
```

---

## 3. Gene Expression Distribution

`plot_expression()` shows how a gene's transcript counts are spread across cell
types. The ggplot2 version overlays a violin (kernel density shape) on a
mini-boxplot for maximum statistical detail.

```{r expr-ggplot, fig.width=7, fig.height=5}
plot_expression(my_data, gene = "CD79A", method = "ggplot2")
```

---

## 4. Descriptive Statistics per Cell Type

`explore_genes()` computes mean, median, and standard deviation for a gene
broken down by cell-type cluster. This reveals whether a gene is a
cluster-specific marker or broadly expressed.

```{r explore}
explore_genes(my_data, gene = "CD79A")
```

---

## 5. Differential Expression — T-test & Wilcoxon

`compare_expression()` runs a two-sample test between any two cell types.
Because scRNA-seq data is often zero-inflated and highly skewed, the
non-parametric Wilcoxon test is popular alongside the classical t-test.

```{r ttest}
# Parametric two-sample t-test
compare_expression(my_data, gene = "CD8A",
                   type1 = "CD8+ T cell", type2 = "B cell",
                   method = "t.test")
```

```{r wilcox}
# Non-parametric Wilcoxon rank-sum test
compare_expression(my_data, gene = "CD8A",
                   type1 = "CD8+ T cell", type2 = "B cell",
                   method = "wilcox")
```

---

## 6. One-Way ANOVA

`run_anova()` tests whether a gene's mean expression differs across **all** cell
types simultaneously. A low p-value indicates the gene is a meaningful
biological marker for at least one cluster.

```{r anova}
result <- run_anova(my_data, gene = "CD8A")
summary(result)
```

---

## Session Info

```{r session}
sessionInfo()
```
